---
title: "Tutorial 12: nMDS, PERMANOVA and SIMPER"
author: "Professor Mathew Crowther"
format:
  sandstone-html:
    code-tools: true
resources:
  - data/macnally.csv
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aim of this tutorial

In the previous tutorial we used **hierarchical clustering** to ask whether sites with similar bird communities group together.

In this tutorial we use the **same Mac Nally bird data**, but analyse it using three related multivariate methods:

1. **nMDS** — to visualise similarity among bird communities.
2. **ANOSIM** — to test whether communities differ among habitat groups.
3. **PERMANOVA** — to test whether habitat type explains multivariate differences in bird assemblages.
4. **SIMPER** — to identify which bird species contribute most to differences among habitat groups.

The data are based on Mac Nally's study of forest and woodland birds along a 250 km transect in south-eastern Australia. The paper examined bird assemblages across contrasting forest and woodland habitats, including river red gum/grey box woodlands, foothill forests, montane forests, red ironbark forests, manna gum woodlands and coastal banksia woodland.

Mac Nally used multivariate approaches to understand habitat structure, bird occurrence and abundance across sites. Here, we focus on the **bird community matrix**: rows are sites and columns are bird species.

# Load packages

```{r}
library(vegan)
library(MASS)
library(ggplot2)
library(dplyr)
```

# Load the data

The dataset contains:

- `SITE` = site name
- `HABITAT` = habitat category
- bird species columns = abundance/density estimates for each bird species

```{r}

Birds <- read.csv("data/macnally.csv", header = TRUE)


head(Birds)
str(Birds)
```

Check the habitat groups:

```{r}
table(Birds$HABITAT)
```

# Create the community matrix

For multivariate community analyses, we need only the species abundance columns.

The first two columns are site information, so the bird species matrix starts at column 3.

```{r}
BirdsData <- Birds[, 3:ncol(Birds)]

head(BirdsData[, 1:8])
dim(BirdsData)
```

Each row is a site.  
Each column is a bird species.

# Transform the data

Community data often contain many zeros and some very abundant species. If we analyse raw abundance data, the ordination may be dominated by only the most abundant species.

A **fourth-root transformation** reduces the influence of very abundant species while still retaining quantitative abundance information.

```{r}
TransBirdsData <- BirdsData^(1/4)
```

# Bray-Curtis dissimilarity

Bray-Curtis dissimilarity is commonly used for ecological community data.

It compares sites based on differences in species composition and abundance.

```{r}
Bird.dis <- vegdist(TransBirdsData, method = "bray")

Bird.dis
```

Interpretation:

- values close to 0 = sites have similar bird communities
- values close to 1 = sites have very different bird communities

# nMDS ordination

nMDS attempts to place sites in a low-dimensional plot so that distances among points reflect their Bray-Curtis dissimilarities.

Sites close together have similar bird assemblages.  
Sites far apart have different bird assemblages.

```{r}
set.seed(123)
nMDS_Bird <- metaMDS(TransBirdsData,
                     distance = "bray",
                     k = 2,
                     trymax = 100)

nMDS_Bird
```

## Stress value

The stress value tells us how well the two-dimensional nMDS represents the multivariate dissimilarities.

As a guide:

- stress < 0.10 = good representation
- stress 0.10–0.20 = usable but interpret with care
- stress > 0.20 = poor representation

# Base R nMDS plot

```{r}
Bird_Factor <- factor(Birds$HABITAT)
pchs <- seq_along(levels(Bird_Factor))

plot(nMDS_Bird,
     display = "sites",
     type = "n",
     main = "nMDS of bird communities")

points(nMDS_Bird,
       display = "sites",
       pch = pchs[Bird_Factor],
       col = "black",
       cex = 1.2)

legend("topright",
       bty = "n",
       legend = levels(Bird_Factor),
       pch = pchs)
```

## Interpretation

If sites from the same habitat type cluster together, this suggests that habitat type is associated with bird community composition.

If habitat groups strongly overlap, this suggests that bird communities are not clearly separated by habitat type.

# ggplot nMDS plot

First extract the nMDS site scores.

```{r}
nmds_scores <- as.data.frame(scores(nMDS_Bird, display = "sites"))
nmds_scores$SITE <- Birds$SITE
nmds_scores$HABITAT <- Birds$HABITAT

head(nmds_scores)
```

Now plot the ordination.

```{r}
ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, shape = HABITAT)) +
  geom_point(size = 3) +
  theme_classic() +
  labs(title = "nMDS ordination of Mac Nally bird communities",
       x = "nMDS axis 1",
       y = "nMDS axis 2",
       shape = "Habitat")
```

# Add habitat ellipses

Ellipses help visualise whether habitat groups occupy different regions of ordination space.

```{r}
ggplot(nmds_scores, aes(x = NMDS1, y = NMDS2, shape = HABITAT)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = HABITAT), type = "t", linetype = 2) +
  theme_classic() +
  labs(title = "nMDS with habitat group ellipses",
       x = "nMDS axis 1",
       y = "nMDS axis 2",
       shape = "Habitat")
```

Note: ellipses are only a visual guide. Formal tests are below.

# ANOSIM

ANOSIM tests whether differences among habitat groups are greater than differences within habitat groups.

```{r}
Bird.anosim <- anosim(Bird.dis, Birds$HABITAT, permutations = 999)

summary(Bird.anosim)
plot(Bird.anosim)
```

## Interpreting ANOSIM

The main statistic is **R**.

- R close to 0 = little separation among groups
- R close to 1 = strong separation among groups
- p-value = whether the observed separation is greater than expected by chance

# PERMANOVA

PERMANOVA tests whether habitat type explains variation in the multivariate bird community data.

In `vegan`, PERMANOVA is run using `adonis2()`.

```{r}
Bird.permanova <- adonis2(TransBirdsData ~ HABITAT,
                          data = Birds,
                          method = "bray",
                          permutations = 999)

Bird.permanova
```

## Interpreting PERMANOVA

Important columns:

- `Df` = degrees of freedom
- `SumOfSqs` = multivariate variation explained
- `R2` = proportion of variation explained by habitat
- `F` = pseudo-F statistic
- `Pr(>F)` = permutation p-value

If the p-value is significant, bird communities differ among habitat types.

# Check dispersion

PERMANOVA can be sensitive to differences in multivariate spread among groups. Therefore, we should check whether habitat groups differ in dispersion.

```{r}
Bird.disp <- betadisper(Bird.dis, Birds$HABITAT)

anova(Bird.disp)
plot(Bird.disp)
```

## Interpretation

If this test is significant, habitat groups differ in their within-group variability.

That does not invalidate PERMANOVA, but it means the PERMANOVA result may reflect differences in spread as well as differences in group centroids.

# Pairwise PERMANOVA comparisons

If the overall PERMANOVA is significant, we may want to know which habitat groups differ from each other.

The package `pairwiseAdonis` is useful, but it may not be installed on all computers. The installation code is shown but not run automatically.

```{r}
#| eval: false
# Install only once, if needed:
# install.packages("devtools")
# devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
```

Load the package if it is already installed:

```{r}
#| eval: false
library(pairwiseAdonis)

pairwise.adonis(TransBirdsData,
                Birds$HABITAT,
                sim.method = "bray",
                p.adjust.m = "bonferroni")
```

If this package is not available, you can still interpret the overall PERMANOVA and the nMDS plot.

# SIMPER analysis

SIMPER stands for **Similarity Percentages**.

It identifies which species contribute most to differences between habitat groups.

```{r}
Birdsim <- simper(TransBirdsData, Birds$HABITAT)

summary(Birdsim)
```

## Extract one comparison

The output can be long. The following code extracts the first pairwise habitat comparison.

```{r}
first_comparison <- summary(Birdsim)[[1]]
head(first_comparison, 10)
```

## Interpretation

Species with larger average contributions are those that most strongly separate two habitat types.

These may be habitat-associated species, but SIMPER should be interpreted carefully because common and abundant species often contribute strongly.

# Questions for students

## Question 1

What does it mean if two sites are close together in the nMDS plot?

### Answer guide

They have similar bird community composition based on Bray-Curtis dissimilarity.

## Question 2

Why did we transform the bird abundance data before calculating Bray-Curtis dissimilarities?

### Answer guide

The transformation reduces the dominance of very abundant species and allows less abundant species to contribute more to the multivariate pattern.

## Question 3

What does a significant PERMANOVA tell us?

### Answer guide

It tells us that bird community composition differs among habitat types more than expected by chance.

## Question 4

Why should we check dispersion with `betadisper()`?

### Answer guide

Because PERMANOVA can be influenced by differences in the spread of groups, not only differences in their centres.

## Question 5

What does SIMPER add to the analysis?

### Answer guide

SIMPER helps identify which species contribute most to the observed differences among habitat types.

# Key take-home message

Hierarchical clustering, nMDS, ANOSIM, PERMANOVA and SIMPER all use the same basic idea: sites can be compared based on their whole community composition.

For the Mac Nally bird data, these tools let us ask whether forest and woodland habitat types support distinct bird assemblages, and which species contribute most to those differences.
