---
title: "Tutorial 07"
# reviewed-by: ACA, 2026
format:
  sandstone-html:
    code-tools: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Welcome to week 7! {.unnumbered}

This week's lecture introduced you to Simple- and Multiple- Linear Regression. For both of these regression techniques we defined the key steps in model fitting and how to interpret model output. 

In this tutorial, we will explore key differences between simple linear regression (SLR) and multiple linear regression (MLR), and we will learn how to fit these models in R. We will also discuss how to test assumptions and evaluate model fit for both types of regression.

You will learn how to:

1.  Distinguish key similarities and differences between simple linear regression (SLR) and multiple linear regression (MLR) models
2.  Implement the fitting process for each type of model
3.  Test assumptions and evaluate model fit of SLR and MLR models.

# 1. Simple linear regression vs Multiple Linear Regression

| Element | Simple Linear Regression (SLR) | Multiple Linear Regression (MLR) |
|:------------------------|:------------------------|:------------------------|
| Number of predictors | 1 | 2 or more |
| Model form | $Y = \beta_0 + \beta_1x + \epsilon$ | $Y = \beta_0 + \beta_1x_1 + + \beta_2x_2 + ... + \beta_nx_n + \epsilon$ |
| How to fit in R | `lm(y ~ x, data = dataset)` | `lm(y ~ x1 + x2 + ... + xn, data = dataset)` |
| Assumptions | L.I.N.E. | L.I.N.E., collinearity |
| Model evaluation | $r^2$, residual plots | Adjusted $r^2$, residual plots, VIF|
| Conclusion | "as $x$ increases, $y$ also increases/decreases by $\beta_1$ units" | "as $x_1$ increases, $y$ also increases/decreases by $\beta_1$ units, holding all other predictors constant" |

: Similarities and differences between MLR and SLR models {#tbl-perf}

# Setting up for our analyses

For both examples we will use the toilet paper dataset (Garbe et al. 2020) which was collected to investigate key factors influencing toilet paper stockpiling during the COVID-19 pandemic. 

A total of 3,648 survey respondents from 22 countries reported their perceived threat of COVID-19, personality traits, and toilet paper consumption behaviour. The study found that perceived threat and Conscientiousness were significant predictors of stockpiling.

Data source: Garbe, Lisa, Richard Rau, and Theo Toppe. 2020a. ŌĆśData: Influence of Perceived Threat of Covid-19 and HEXACO Personality Traits on Toilet Paper StockpilingŌĆÖ. OSF. https://osf.io/nbrg5/.

There was a study that investigated key predictors of toilet paper stockpiling d
 Survey respondents from 22 countries reported their perceived threat of COVID-19, personality traits, and toilet paper consumption behaviour. The study found that perceived threat and Conscientiousness were significant predictors of stockpiling.

# 2. Implement the fitting process for each type of model



# 3. Test assumptions and evaluate model fit of SLR and MLR models.









## Bird abundance

Fragmentation of forest habitat has an impact of wildlife abundance. This study looked at the relationship between bird abundance (bird ha^-1^) and the characteristics of forest patches at 56 locations in SE Victoria.

The predictor variables are:

-   `ALT` Altitude (m)
-   `YR.ISOL` Year when the patch was isolated (years)
-   `GRAZE` Grazing (coded 1-5 which is light to heavy)
-   `AREA` Patch area (ha)
-   `DIST` Distance to nearest patch (km)
-   `LDIST` Distance to largest patch (km)

Import the data from the "Loyn" tab in the MS Excel file.

```{r}
library(readxl)
loyn <- read_xlsx("data/mlr.xlsx", "Loyn")
```

Often, the first step in model development is to examine the data. This is a good way to get a feel for the data and to identify any issues that may need to be addressed. In this case, we will examine the data using histograms and a correlation matrix.

### Histograms

There are a breadth of ways to create histograms in R. In each tab below you will find some different ways to create the same plot outputs.

::: panel-tabset
### `hist()`

This is a straightforward way to create multiple histograms with `hist()`. The `par()` function is used to arrange the plots on the page. The `mfrow` argument specifies the number of rows and columns of plots.

```{r}
# par(mfrow=c(3,3))
hist(loyn$ABUND)
hist(loyn$ALT)
hist(loyn$YR.ISOL)
hist(loyn$GRAZE)
hist(loyn$AREA)
hist(loyn$DIST)
hist(loyn$LDIST)
# par(mfrow=c(1,1))
```

### `hist.data.frame()` from `Hmisc`

The `Hmisc` package provides a function `hist.data.frame()` that can be used to create multiple histograms, which can be called by simply using `hist()`. You may need to tweak the `nclass` argument to get the desired number of bins, as the default may not look appropriate.

```{r eval=FALSE}
#| message: false
# install.packages("Hmisc")
library(Hmisc)
hist(loyn, nclass = 50)
```

### `ggplot()`

A more modern approach is to use `ggplot()` with `facet_wrap()` to arrange multiple plots on a single page. To do this, the `pivot_longer()` function from the `tidyr` package is used to reshape the data into a tidy format.

```{r}
#| message: false
# tidy the data
loyn_tidy <- pivot_longer(loyn, cols = everything())

# plot
ggplot(loyn_tidy, aes(x = value)) +
    geom_histogram() +
    facet_wrap(~name, scales = "free") +
    theme_bw()
```

### `ggplot()` with `dplyr`

Here we use the pipe operator `%>%` from `dplyr` to chain together a series of commands. The pipe operator takes the output of the command on the left and passes it to the command on the right (or below) the pipe. This means that we can create a series of commands that are executed in order.

```{r}
#| message: false
loyn %>%
    pivot_longer(cols = everything()) %>%
    ggplot(aes(x = value)) +
    geom_histogram() +
    facet_wrap(~name, scales = "free") +
    theme_bw()
```
:::

::: question
### Question 1

Comment on the histograms in terms of leverage. *Hint: what is the relationship between leverage and skewness?*
:::

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

The histograms of `AREA`, `DIST` and `LDIST` are very skewed. The high values would have high leverage, this means that these would cause the residuals to be skewed. These would be candidates for transformation.
:::

### Correlation matrix

Calculate the correlation matrix using `cor(Loyn)`.

```{r}
cor(loyn)
```

::: question
### Question 2

Which independent variables are useful for predicting the dependent variable abundance? Is there evidence for multi-collinearity?
:::

```{r corLoyn, eval=F, echo=F}
cor(loyn)
```

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

Some of the predictors are useful, but AREA has a low r.

The correlation between `GRAZE` and `YR.ISOL` is quite high (r = -0.63556710), suggesting multi-collinearity which may influence the model. If the relationship between these two variables was stronger, we would remove one of the variables to prevent this collinearity from affecting the model.

*Note: For more information on collinearity and how it may impact the model, see Quinn & Keough p 127.*
:::

### Plotting correlation

Examine correlations visually using `pairs()` or `corrplot()` from the `corrplot` package.

::: panel-tabset
### Scatterplot matrix

```{r}
pairs(loyn)
```

### Correlation matrix

```{r}
#| message: false
library(corrplot)
corrplot(cor(loyn))
```
:::

::: question
### Question 3

Are there any trends visible from the plots?
:::

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

Not really; the pairs plot reflects the strength of the linear relationship between each of the variables. There may be some stronger relationships occurring, but it is evident a few of the variables are skewed so it is harder to distinguish within the plots.
:::

::: callout-tip
We can also bring in variance inflation factors (VIF) to help us identify multi-collinearity, but that is done only after we have selected a model.
:::

### Transformations

The AREA predictor has a small number of observations with very large values. Apply a log~10~ transformation and label the new variable `Loyn$L10AREA`.

```{r}
#| eval: false
loyn$L10AREA <- log10(loyn$AREA)
```

::: question
### Question 4

Why are we transforming AREA?
:::

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

You do this to stabilise the variance of the regression to manage the leverage of the outliers in the variable. This reduces the skew. L10AREA is more likely to be a significant predictor.
:::

::: question
### Question 5

Re-run `pairs(Loyn)` and create a histogram using the transformed value of AREA, how do the plots look?
:::

```{r}
#| eval: false
hist(loyn$L10AREA)
pairs(loyn)
```

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

-   Histogram looks better, less skewed
-   Pairs plot shows a trend between ABUND and L10AREA
:::

::: question
### Question 6

In preparation for modelling, transform the remaining skewed variables, DIST and LDIST the same way you did for AREA and examine the histogram and pairs plots using these new variables.
:::

Make sure you end up with two new variables labelled `loyn$L10DIST` and `loyn$L10LDIST`.

```{r}
#| eval: false
#| include: false
loyn$L10DIST <- log10(loyn$DIST)
hist(loyn$L10DIST)

loyn$L10LDIST <- log10(loyn$LDIST)
hist(loyn$L10LDIST)

pairs(loyn)
```

::: {.content-visible when-profile="solution"}
### Answer {style="color:green;"}

Histogram for both look better, less skewed Pairs plot shows potential trend between ABUND and L10DIST, and ABUND with L10LDIST compared to untransformed DIST and LDIST
:::

::: callout-note
## Checkpoint

You should now be able to ...\[PUT THIS AFTER EACH OBJECTIVE\]
:::

## Wrap-up {.unnumbered}

In this tutorial, **we explored factorial designs and their analysis using ANOVA. We learned about main effects, interactions, and blocking effects, and how to interpret ANOVA results in the context of factorial experiments. We also visualised these effects using ggplot2 and emmeans, enhancing our understanding of the relationships between factors and response variables in experimental designs.**

### Example exam questions

**1. In a 2-way factorial design with factors A (2 levels) and B (3 levels), describe how you would identify and interpret main effects and interactions using ANOVA. Provide an example of how to visualise these effects.**

2.  Explain the difference between a treatment design and an experimental design. Provide examples of each and discuss how they can be combined in a factorial design with blocking.

3.  **If you conducted a factorial experiment with blocking and found a significant interaction between two factors, how would you interpret this result? Would you still consider the main effects of each factor? Justify your answer with reference to ANOVA principles.**
