---
title: Lab 09
# revised-by: aca + Liana
subtitle: ENVX2001 Applied Statistical Methods
description: The University of Sydney
toc: true
toc-depth: 3
resources:
  - data/Loyn_SLR.xlsx
  - data/california_streamflow.xlsx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Welcome to week 9 - the last week of modelling!

This week's lab will give you some practice testing how well our models predict new values.  

Next week you will start on multivariate approaches.  

## Learning outcomes

This module links to the following unit learning outcomes: 

* LO3. demonstrate proficiency in the use or the statistical programming language R to apply an ANOVA and fit regression models to experimental data

* LO5. interpret the output and understand conceptually how its derived of a regression, ANOVA and multivariate analysis that have been calculated by R

* LO6. write statistical and modelling results as part of a scientific report

In this lab, we will learn how to:

**1. Use the parsimonious models we produced last week to predict new values**

**2. Interpret key prediction quality metrics - do our models model predict well?**


## Specific goals

**By the end of this lab, you should be able to:**

- [ ] Fit a multiple linear regression model
- [ ] Interpret VIF plots and output to determine whether collinearity is a concern in your model
- [ ] Remove non-significant predictors to produce a more parsimonious model
- [ ] Compare a full and reduced model using an F-test
- [ ] Run and interpret stepwise regression and associated output
- [ ] Compare a full and reduced model using the AIC
- [ ] Use the results of model comparisons to select the best model for your data

## Preparation

Please work on these exercises by creating your own Quarto file. You are welcome to write your responses up in the style of a report.

### Packages

This lab uses `tidyverse`, `readxl`, `moments`, `psych`, `performance`, `car`. Install any you are missing by running the following **in the console**:

```{r}
#| eval: false

install.packages(c("car", "tidyverse", "readxl", "moments", "psych", "performance"))
```

If a package is already installed, R will simply reinstall it — no harm done.

# 1. Variable selection for California streamflow (~ 45 mins)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Learning Outcomes

In this lab, you will work towards achieving learning outcomes

## Lab Objectives

In this lab, we will:

- []
- []


:::{.callout-tip}
Please work on this exercise by creating your own R Markdown file.
:::

## Exercise 1:

## Exercise 2:

## Take home exercise:

## Review

### Attribution 

