---
title: Lecture 03a -- $t$-Tests
subtitle: ENVX2001 Applied Statistical Methods
date: last-modified
date-format: "MMM YYYY"
author: Januar Harianto
format:
  ochre-revealjs: 
    slide-number: c/t
  ochre-typst:
    fig-format: png
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
```

# The $t$-test (revision)

## Introduction

- **One-sample** $t$-test: compare a sample mean against a known value
- **Two-sample** $t$-test: compare the means of two independent groups
- **Paired** $t$-test: compare measurements taken on the same subjects (e.g. before and after)

Last week we learned about **confidence intervals**, which estimate the uncertainty of a parameter (e.g. a mean).

::: fragment
The $t$-test and the confidence interval are closely related. Both are derived from the same quantities, and this week we look at that connection more carefully.
:::


## Cattle

Two breeds of cattle, 12 and 15 samples. Are the breeds different in mean weight?

```{r}
#| code-fold: true
#| results: hide
#| fig-align: center
cattle <- read.csv("data/cattle.csv") |>
  pivot_longer(cols = everything(), names_to = "breed", values_to = "weight") |>
  mutate(breed = as.factor(breed)) |>
  drop_na()

ggplot(cattle, aes(breed, weight)) +
  geom_boxplot() +
  labs(x = "Breed", y = "Weight (kg)") + 
  cowplot::theme_cowplot()
```

::: fragment
The boxplot suggests a difference. To test, you would run `t.test()` and check the p-value. **What is the test actually computing?**
:::


# From CIs to the $t$-test

## The difference between two means

In the last lecture, we constructed confidence intervals for a single mean. The same principle extends to the **difference** between two means.

- Estimate the difference: $\bar{y}_1 - \bar{y}_2$
- Construct a 95% confidence interval around that estimate
- Evaluate whether the interval includes zero

::: fragment
- If the interval **excludes zero**: the means are significantly different at the 5% level
- If the interval **includes zero**: the data are consistent with no difference
:::


## `t.test()` gives you both

```{r}
#| code-fold: true
t.test(weight ~ breed, data = cattle, var.equal = TRUE)
```

The output reports two quantities:

- A **p-value** (probability of observing a result this extreme if the null hypothesis were true)
- A **95% confidence interval** for the difference between the two means

::: fragment
These always agree. If the CI excludes zero, the p-value is below 0.05, and vice versa. The $t$-statistic measures how far the observed difference is from zero, relative to its standard error.
:::


## Exploring the confidence interval

```{=html}
<style>
#exploring-the-confidence-interval .cell-output-display {
  border: none;
  padding: 0;
  margin: 0;
}
</style>
```

```{ojs}
//| echo: false
import { ciDifferenceWidget } from "../../assets/js/ttest-widget.js"
ciDifferenceWidget()
```

- Increasing the difference between means moves the CI away from zero
- Decreasing variability narrows the CI
- When the interval excludes zero, the means are significantly different


## The CI for a difference

In the last lecture, we built a 95% CI for a single mean:

$$\bar{y} \pm t^* \times SE$$

::: fragment
For the **difference** between two means, the same structure applies:

$$(\bar{y}_1 - \bar{y}_2) \pm t^* \times SE(\bar{y}_1 - \bar{y}_2)$$

If this interval excludes zero, the groups differ significantly.
:::


## The $t$-statistic

We can rearrange the CI formula to isolate a single number that captures both the difference and the variability:

$$t = \frac{\bar{y}_1 - \bar{y}_2}{SE(\bar{y}_1 - \bar{y}_2)}$$

The $t$-statistic counts how many standard errors the observed difference is from zero.

::: fragment
For the cattle data: $t = 42.5 / 4.5 = 9.46$. The observed difference is over nine standard errors from zero.
:::


## From the $t$-statistic to the p-value

We have a $t$-statistic. The p-value answers: if there were truly no difference between the breeds, how often would we see a value of $|t|$ this large or larger?

- A small p-value means the observed difference is hard to explain by chance alone
- A large p-value means the observed difference is consistent with random variation

::: fragment
The 95% CI and the p-value always agree. If the 95% CI excludes zero, the p-value is below 0.05, and vice versa.
:::


## Exploring the p-value

```{=html}
<style>
#exploring-the-p-value .cell-output-display {
  border: none;
  padding: 0;
  margin: 0;
}
</style>
```

```{ojs}
//| echo: false
import { pvalueExplorerWidget } from "../../assets/js/ttest-widget.js"
pvalueExplorerWidget()
```

- The shaded area under both tails is the p-value
- As the $t$-statistic moves further from zero, the p-value decreases
- At low degrees of freedom the tails are heavier, but the shape approaches the normal as df increases


# The hypothesis testing workflow

## Step 1: State what we are testing

- **Null hypothesis**: $H_0: \mu_1 = \mu_2$ (no difference in mean weight between breeds)
- **Alternative hypothesis**: $H_1: \mu_1 \neq \mu_2$ (means differ)
- This is a **two-tailed** test: we are looking for a difference in either direction (Breed 1 heavier *or* lighter than Breed 2), not just one

::: fragment
**Model equation**: $y_{ij} = \mu_i + \varepsilon_{ij}$

Each observation is the group mean plus random error. The hypothesis asks whether $\mu_1$ and $\mu_2$ are equal.
:::


## Step 2a: Equal variances

The $t$-test assumes both groups have similar variability. If they do not, then the test is unreliable because the pooled estimate of variance will not represent either group well.

- Rule of thumb: the ratio of the larger SD to the smaller SD should be below 2.0

```{r}
#| code-fold: true
cattle_summary <- cattle |>
  group_by(breed) |>
  summarise(n = n(), mean_wt = mean(weight), sd_wt = sd(weight))

cattle_summary
max(cattle_summary$sd_wt) / min(cattle_summary$sd_wt)
```


## Step 2b: Normality

The $t$-test assumes the data in each group follow a normal distribution. With small samples, departures from normality (strong skew, heavy tails, outliers) can distort the p-value and confidence interval. Check each group separately.

```{r}
#| code-fold: true
shapiro.test(cattle$weight[cattle$breed == "Breed1"])
shapiro.test(cattle$weight[cattle$breed == "Breed2"])
```


## Step 2b: Normality (QQ plots)

```{r}
#| code-fold: true
#| fig-align: center
#| layout-ncol: 2
qqnorm(cattle$weight[cattle$breed == "Breed1"], main = "Breed 1")
qqline(cattle$weight[cattle$breed == "Breed1"])
qqnorm(cattle$weight[cattle$breed == "Breed2"], main = "Breed 2")
qqline(cattle$weight[cattle$breed == "Breed2"])
```

If points fall close to the line, the data are approximately normal. Both groups pass the Shapiro-Wilk test ($p > 0.05$) and the QQ plots show no strong departures.


## Step 2c: Independence

Observations should not influence each other. This is a property of the study design, not something we test statistically.

- Each animal is measured once
- The weight of one animal does not affect the weight of another
- If the same animals were measured before and after a treatment, observations would not be independent (use a paired $t$-test instead)


## Step 3: Run the test and interpret

```{r}
t.test(weight ~ breed, data = cattle, var.equal = TRUE)
```

- The $t$-statistic, degrees of freedom ($n_1 + n_2 - 2$), and p-value are reported together
- `var.equal = TRUE` performs the pooled (Student's) $t$-test, which assumes equal variances

::: fragment
A p-value of 0.03 does not mean there is a 3% chance the null hypothesis is true. It means: if there were truly no difference, we would observe a result this extreme about 3% of the time.
:::


## HATPC

Notice the five steps we just followed:

1. **H**ypotheses: state the null and alternative
2. **A**ssumptions: check equal variances, normality, independence
3. **T**est: run the appropriate test
4. **P**-value: interpret the result
5. **C**onclusion: state the finding in context

::: fragment
We will use this framework throughout the semester, starting with ANOVA in the next lecture.
:::


## What if assumptions are not met?

::: incremental
- **Unequal variances**: use Welch's $t$-test (the default in R when `var.equal = FALSE`)
- **Non-normality**: if $N > 30$, the Central Limit Theorem provides reasonable coverage
- **Non-independence**: consider a **paired $t$-test** (e.g. before and after measurements on the same subjects)
:::


# Summary

## Key points

- The **two-sample $t$-test** compares means of two independent groups
- The $t$-test and the CI for the difference are two views of the same calculation
- Check the three assumptions (equal variances, normality, independence) before running any test
- Follow **HATPC** as a structured process for hypothesis testing

::: fragment
In Lab 03, you will practise this workflow with `t.test()` and diagnostic plots.
:::

::: fragment
**Next**: what if we have more than two groups? We need **ANOVA**.
:::


## 15 minute break {.center-h}

This presentation is based on the [SOLES Quarto reveal.js template][soles-revealjs] and is licensed under a [Creative Commons Attribution 4.0 International License][cc-by].


[soles-revealjs]: https://github.com/usyd-soles-edu/soles-revealjs
[cc-by]: http://creativecommons.org/licenses/by/4.0/
